\documentclass[../../thesis.tex]{subfiles}

\begin{document}

This thesis is the culmination of 6 years of studies in Trondheim. There are many people that deserve big thank you for making these years so memorable. If you are reading this, chances are quite high that your name should be here.\newline 

I would firstly like to thank my supervisors, Erlend Aune and Daesoo Lee, for introducing me to this fascinating research field, and being open minded and supportive throughout the process. Erlend Lokna, a dear friend and great collaborator, for your indispensable contributions on this project. Developing advanced machine learning models is both challenging and time consuming, and synchronously banging our heads made it all a little easier. From the many late nights of writing, experimentation and discussion, to trash talking over the ping pong table, this project would not be the same without you. It was sweet to end on a 9-1 victory, which once and for all cemented me as the top dog, and nothing (except a rematch) could change that.\newline

I would too like to thank my family, mom, dad, Otto, Andreas and Tiril for being a continuous source of inspiration and motivation. You move so graciously through life, sharing both of wisdom and skill, paving the way for the youngest in the flock. Erik, Ludvig and Sivert, my day ones. You have been there through all the ups and downs, with unwaivering support just a small phone call away. My roommates, Pernille and Eldrun, for being devilishly funny and handling me so well as I go full goblin mode. Preben, Elias and the rest of the Matteland mafia, despite placing bets on whether I would last one or two weeks into my bachelors, for their endless curiosity and fearlessness in face of difficult mathematics, pushing me to learn way more than I aught to have. \newline

Finally, Tindegruppa, the university climbing group, my crew. You have such a special place in my heart. You reintroduced the joy of movement to my life, after abruptly retiring as professional handball player. You provided an arena to play, connect with nature and explore. Someone once told me, to have a great time as a student in Trondheim, you need to find your little community, be it on Samfundet or some part of NTNUI. I can wholeheartedly say I found mine, and to all my climber friends, such loving, crazy and interesting people, thank you. See you on top of some remote peak very soon.

\end{document}



